---
title: "Where pcdreg fits among R packages for panel count data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Where pcdreg fits among R packages for panel count data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5)
```

```{r setup}
library(pcdreg)
```

This article is about choosing a tool, so it says what `pcdreg` does that the
alternatives do not, and equally what they do that it does not.

# The landscape

Three kinds of package meet recurrent event data in R, and they are not
interchangeable.

**Exactly observed event times.** If you know *when* each event happened,
`reReg`, `survival` and the rest of the recurrent event literature apply, and
none of what follows is relevant. Panel count data are the case where you know
only how many events fell between visits.

**`spef`**, by Chiou, Wang and Yan, is the established R package for panel count
data. It offers a wide range of estimators through one function,
`panelReg(formula, data, method = ...)`: augmented estimating equations that
allow the observation process to be informative (`AEE`, `AEEX`), maximum
pseudolikelihood and monotone spline likelihood methods (`MPL`, `MPLs`, `MLs`),
the Sun--Wei and Hu--Sun--Wei estimating equations, and an accelerated mean
model. Its response is `PanelSurv(ID, time, count)`, one row per examination.
Every one of those methods targets the **mean** model. At the time of writing
`spef` is archived on CRAN — since June 2026, for an undeliverable maintainer
address rather than anything to do with the code — so it installs from the
archive rather than from `install.packages()`.

**`pcdreg`** fits the rate model and the mean model, with covariates that may
vary over time, and estimates the covariance without leaning on a Poisson
assumption.

# What is actually different

## The rate model

This is the substantive difference. The means model
\eqn{E[N(t)\mid X(t)] = \mu(t)\exp(\beta'X(t))} constrains the *cumulative*
count; the rate model \eqn{E[dN(t)\mid X(t)] = \exp(\beta'X(t))d\Lambda(t)}
constrains the *instantaneous* one. With covariates that move over time these
are different models, not two parametrisations of one, and \eqn{\beta} means
something different in each.

Simulate from the rate model with a covariate that steps part way through
follow-up, and fit both:

```{r bothmodels}
set.seed(1)
d <- r_panel_count(300, beta = c(1, -1), lambda = function(t) 8 / (1 + t))

rate <- pcdreg(pcd(id, tstart, tstop, count) ~ x1 + x2, data = d)
mean <- pcdreg(pcd(id, tstart, tstop, count) ~ x1 + x2, data = d,
               model = "mean")

round(cbind(truth = c(1, -1), rate = coef(rate), mean = coef(mean)), 3)
```

The rate model recovers the values used to generate the data. The means model
does not, and it is not meant to: it is estimating a different quantity. If your
scientific question is about the instantaneous risk — the analogue of a hazard
ratio — then only the first column of estimates answers it, and a package
offering mean models alone cannot produce it.

The reverse holds too. If your question is about the cumulative burden by time
\eqn{t}, the means model is the right one, and `pcdreg(model = "mean")` and
`spef`'s `EE.HSWc` are the same estimator.

## Monotonicity

There is a second, more practical consequence. Nothing constrains a fitted mean
function to increase, and with a fluctuating covariate it generally does not:

```{r monotone}
mu <- baseline(mean)$mean
c(increasing = !is.unsorted(mu), decreases_at = sum(diff(mu) < 0))
```

A mean function that falls is not a numerical failure; it is what the estimator
returns, and it makes the fitted model awkward to interpret and to predict from.
The rate model has no such problem, because it constrains only
\eqn{\exp(\beta'X(t))d\Lambda(t)} to be positive, so predicted means increase by
construction:

```{r predmono}
pr <- predict(rate, d, type = "mean")
all(vapply(split(pr$mean, pr$id), function(v) all(diff(v) >= 0), logical(1)))
```

Some `spef` methods address this from the other side, fitting \eqn{\mu} with
monotone I-splines so the *baseline* increases by construction. That is a real
solution to half the problem: it does not stop the fitted mean for a given
subject from falling when their covariates fall.

## Covariates between examinations

The rate model's likelihood evaluates each subject's covariate trajectory at
every pooled examination time inside their follow-up, including times at which
*that* subject was not examined. A response of the form
`PanelSurv(ID, time, count)`, with one row per examination, has nowhere to put a
covariate value at a time when the subject was not seen. `pcd()`'s counting
process form does:

```{r layout}
head(subset(d, id == 1), 4)
```

Row 3 here has `count = NA`: it records that a covariate changed at that time,
not an examination. That is the information the rate model needs and the one row
per examination layout cannot carry.

## Variance without the Poisson assumption

The rate model's likelihood is the one a Poisson process would give, which is a
working device rather than a claim about the data. Recurrent event data are
usually overdispersed, and then the standard errors that assume Poisson counts
are far too small. `pcdreg` reports a robust sandwich by default and will show
you the difference:

```{r robust}
set.seed(2)
od <- r_panel_count(300, beta = c(1, -1), frailty = 1)  # gamma frailty
odfit <- pcdreg(pcd(id, tstart, tstop, count) ~ x1 + x2, data = od)

round(rbind(robust      = sqrt(diag(vcov(odfit, "robust"))),
            information = sqrt(diag(vcov(odfit, "information")))), 4)
```

The simulations in the paper put the coverage of nominal 95% intervals built
from the second row at 15--22%. Both are available here, and the honest one is
the default.

# Seeing the data first

`pcdreg` also plots the observation pattern, which is worth a look before
fitting anything:

```{r dataplot, fig.alt = "Panel count data plot: one row per subject, follow-up drawn as a grey line, filled circles at examinations with area proportional to the event count and open circles where no events were recorded."}
plot(with(d, pcd(id, tstart, tstop, count)), max_subjects = 40)
```

Two things visible here decide how the analysis will go. Whether examination
times line up across subjects sets the size of the pooled grid and so the cost
of fitting the rate model; and how ragged the right hand edge is tells you how
much of the tail of the baseline rests on a handful of subjects.

# Which to use

| If you need | Use |
|---|---|
| the rate model, or covariates that change between visits | `pcdreg` |
| standard errors that survive overdispersion | `pcdreg` |
| the mean model with the Hu--Sun--Wei estimating equation | either |
| informative observation or censoring times | `spef` (`AEE`, `AEEX`) |
| a monotone spline baseline mean | `spef` (`MPLs`, `MLs`) |
| an accelerated mean model, or the other mean-model estimators | `spef` |
| exactly observed event times | not panel count data; see `reReg` |

`pcdreg` is deliberately narrow. It implements two models carefully, with the
covariance estimation worked out and tested, rather than many models at once.
Where `spef` has an estimator you need, use it.

# Reference

Sun, D., Guo, Y., Li, Y., Tu, W. and Sun, J. (2024). A robust approach for
regression analysis of panel count data. *Bernoulli* **30**(4), 3251--3275.
[doi:10.3150/23-BEJ1713](https://doi.org/10.3150/23-BEJ1713)
