---
title: "Preparing data and specifying the model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing data and specifying the model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(pcdreg)
```

Getting the data into the right shape is most of the work of fitting these
models, and it is where the mistakes are. This vignette covers what the two
fitters expect, how to say it in a formula, how to convert from the layouts data
usually arrives in, and what each error message means.

# What panel count data is

A subject is examined at times $T_{i1} < \dots < T_{iJ_i}$. At each examination
you learn how many events have happened since the previous one, but not when any
of them happened. Two quantities matter and are easy to confuse:

* $\Delta N_{ij}$, the **increment** — events in $(T_{i,j-1}, T_{ij}]$;
* $N_i(T_{ij})$, the **cumulative** count — events up to $T_{ij}$.

**`pcdreg` always wants the increment.** If your data records cumulative counts,
difference them first; there is a recipe below. Passing cumulative counts is not
an error the package can detect, and it will quietly give you the wrong answer.

# The response

`pcd()` builds the left hand side of the formula, in the same way `Surv()`
does for survival models. It has two forms, and which you need depends on one
question only: **do any covariates change value during follow-up?**

## No time-varying covariates: one row per examination

```{r three-arg}
visits <- data.frame(
  id    = c(1, 1, 1, 2, 2),
  time  = c(0.5, 1.1, 1.9, 0.7, 1.4),
  count = c(2, 0, 3, 1, 4),
  age   = c(24, 24, 24, 31, 31),
  arm   = c("drug", "drug", "drug", "placebo", "placebo")
)

pcd(visits$id, visits$time, visits$count)
```

Interval starts are filled in from each subject's previous examination, with
follow-up beginning at zero, which is what the printed intervals show. Subject 1
contributes $(0, 0.5]$, $(0.5, 1.1]$ and $(1.1, 1.9]$ with 2, 0 and 3 events.
Fitting is then

```{r three-arg-fit, eval = FALSE}
pcdreg(pcd(id, time, count) ~ age + arm, data = visits)
```

though not on five rows and two subjects: `age` and `arm` are perfectly
confounded there, and the information matrix is singular. Real fits appear later
in this vignette.

## Time-varying covariates: counting process rows

When a covariate changes between examinations, you need a row for each stretch
over which the covariates are constant. This is exactly the layout `coxph()`
uses for time-dependent covariates.

```{r four-arg}
cp <- data.frame(
  id     = c(1, 1, 1, 2),
  tstart = c(0.0, 0.5, 0.9, 0.0),
  tstop  = c(0.5, 0.9, 1.2, 0.8),
  count  = c(2, NA, 0, 3),
  dose   = c(10, 10, 25, 25)
)

pcd(cp$id, cp$tstart, cp$tstop, cp$count)
```

Read it as: subject 1 was examined at 0.5 with 2 events and at 1.2 with none,
and their dose changed from 10 to 25 at time 0.9, part way through the second
examination interval.

The rule for `count` is what distinguishes the two kinds of row:

* a **number** means "this row ends at an examination, and this many events
  happened since the previous examination";
* **`NA`** means "this row ends at a covariate change, not an examination".

That is the whole convention. There is no separate event indicator.

## Which form, in one line

Use the three argument form when every covariate is fixed at baseline. Use the
four argument form as soon as one covariate moves. If in doubt use the four
argument form: it can express anything the three argument form can.

# The formula

The right hand side is an ordinary model formula, so the usual things work.

```{r rhs, eval = FALSE}
pcd(id, tstart, tstop, count) ~ x1 + x2          # main effects
pcd(id, tstart, tstop, count) ~ x1 * x2          # interaction
pcd(id, tstart, tstop, count) ~ x1 + log(x2)     # transformation
pcd(id, tstart, tstop, count) ~ x1 + factor(site)
pcd(id, tstart, tstop, count) ~ .                # everything else
```

Factors are expanded with your contrast settings, and `predict()` reuses the
coding the fit used.

```{r factors}
set.seed(1)
d <- r_panel_count(60)
d$site <- factor(c("north", "south", "east")[1 + d$id %% 3])

fit <- pcdreg(pcd(id, tstart, tstop, count) ~ x1 + site, data = d)
names(coef(fit))
```

The dot expands to every column that is not part of the response, so `id`,
`tstart`, `tstop` and `count` are excluded automatically:

```{r dot}
names(coef(pcdreg(pcd(id, tstart, tstop, count) ~ ., data = d)))
```

## There is no intercept

A constant column cannot be told apart from a rescaling of the baseline
function, so both fitters drop the intercept and report the baseline separately
through `baseline()`. Writing `~ x1 + x2 - 1` changes nothing, and adding `+ 0`
is unnecessary.

One consequence worth knowing: adding a constant $c$ to a covariate leaves
$\beta$ untouched and rescales the baseline by $e^{-c\beta}$. Centring your
covariates changes the baseline, not the coefficients.

## Row order does not matter

The data are sorted internally by subject and time, so you do not need to
`arrange()` first. Nor do subject identifiers need to be integers or
consecutive: numbers, characters and factors all work, and the original labels
come back in `predict(type = "mean")`.

# Recipes for common layouts

## Cumulative counts

The most common and most damaging mismatch. If your `count` column accumulates,
difference it within subject before fitting.

```{r cumulative}
# A frame whose counts accumulate: 2, 2, 5 rather than 2, 0, 3.
cumdat <- data.frame(
  id    = c(1, 1, 1, 2, 2),
  time  = c(0.5, 1.1, 1.9, 0.7, 1.4),
  count = c(2, 2, 5, 1, 5),
  age   = c(24, 24, 24, 31, 31)
)

cumdat$increment <- ave(cumdat$count, cumdat$id,
                        FUN = function(v) v - c(0, v[-length(v)]))
cumdat[, c("id", "time", "count", "increment")]
```

Then fit with `increment`. With `NA`s present for covariate-change rows, do the
differencing on the examination rows alone:

```{r cumulative-cp, eval = FALSE}
at <- !is.na(dat$count)
dat$count[at] <- ave(dat$count[at], dat$id[at],
                     FUN = function(v) v - c(0, v[-length(v)]))
```

## Data held in the older `panelEM` layout

That layout repeated each panel count on every row of its interval and marked
examinations with a separate indicator. A repeated count carries nothing the
count on the examination row does not, so the conversion is one line:

```{r legacy, eval = FALSE}
new <- old
new$count[old$event == 0] <- NA
pcdreg(pcd(id, tstart, tstop, count) ~ x1 + x2, data = new)
```

## Examinations and covariate histories in separate tables

If visits and covariate changes live in different tables, stack them and carry
the covariates forward. The examination rows keep their counts; the covariate
change rows get `NA`.

```{r merge}
exams <- data.frame(id = c(1, 1), time = c(0.5, 1.2), count = c(2, 0))
covar <- data.frame(id = 1, time = 0.9, dose = 25)
start <- data.frame(id = 1, time = 0, dose = 10)

both <- merge(
  data.frame(id = c(exams$id, covar$id), tstop = c(exams$time, covar$time),
             count = c(exams$count, NA)),
  rbind(start[, c("id", "time", "dose")], covar[, c("id", "time", "dose")]),
  by.x = c("id", "tstop"), by.y = c("id", "time"), all.x = TRUE
)
both <- both[order(both$id, both$tstop), ]
both$dose <- ave(both$dose, both$id,
                 FUN = function(v) { for (i in seq_along(v))
                   if (is.na(v[i]) && i > 1) v[i] <- v[i - 1]; v })
both$tstart <- ave(both$tstop, both$id,
                   FUN = function(v) c(0, v[-length(v)]))
both[, c("id", "tstart", "tstop", "count", "dose")]
```

## One row per subject, several visit columns

Wide data needs reshaping to long first; `stats::reshape()` or
`tidyr::pivot_longer()` both do it. Aim for one row per subject and visit, then
use the three argument form.

# What the data must satisfy

Both fitters check these before doing any arithmetic, and name the offending
subject when a check fails.

**Follow-up starts at zero.** The models integrate from 0, so the first interval
of every subject must begin there.

```{r err-origin, error = TRUE}
bad <- data.frame(id = 1, tstart = 0.2, tstop = 1, count = 3, x = 1)
pcdreg(pcd(id, tstart, tstop, count) ~ x, data = bad)
```

If your times are calendar dates, or follow-up starts at enrolment, subtract each
subject's own origin first.

**Intervals are contiguous.** Within a subject each interval must begin exactly
where the previous one ended: no gaps, no overlaps.

```{r err-gap, error = TRUE}
bad <- data.frame(id = c(1, 1), tstart = c(0, 0.7), tstop = c(0.5, 1),
                  count = c(2, 1), x = c(1, 1))
pcdreg(pcd(id, tstart, tstop, count) ~ x, data = bad)
```

A gap usually means a covariate history that does not cover the whole of
follow-up. Carry the last value forward rather than leaving the stretch out.

**Every subject has at least one examination.** A subject whose `count` is `NA`
on every row contributes nothing and is more likely a data error than an
intention.

**Counts are non-negative whole numbers.** `pcd()` refuses anything else,
which catches rates and averages passed by mistake.

```{r err-count, error = TRUE}
pcd(1:2, c(1, 2), c(0.5, 1.2))
```

**One row per subject may end at any given time.** Duplicate rows for the same
subject and time are rejected rather than silently combined.

## What gets dropped

Follow-up ends at a subject's last examination: the likelihood runs over
$j = 1, \dots, J_i$ and knows nothing beyond $T_{iJ_i}$. Rows lying entirely
after it are dropped, with a message saying how many.

```{r dropped}
# A censoring row appended after subject 1's final visit.
last1 <- max(d$tstop[d$id == 1])
trailing <- transform(d[d$id == 1, ][1, ],
                      tstart = last1, tstop = last1 + 0.5, count = NA)

fit2 <- pcdreg(pcd(id, tstart, tstop, count) ~ x1 + x2,
                  data = rbind(d, trailing))
```

If that message surprises you, the usual cause is a censoring row appended after
the last visit. It is harmless: no information is lost, because there was none in
that stretch to begin with.

## Missing values

Missing covariates are handled by `na.action`, which by default deletes rows —
and deleting a row in the middle of a subject breaks contiguity, so you will get
the gap error rather than a silently different fit. Resolve missing covariates
before fitting, by carrying values forward or by dropping whole subjects.

# What the package does with your rows

Knowing this makes the error messages obvious.

The pooled grid $t_1 < \dots < t_K$ is the set of all distinct examination times
across all subjects. For the **rate model**, each subject's covariate trajectory
is evaluated at every grid time inside their follow-up, which is why the whole
trajectory must be present and contiguous. For the **means model** only the
subject's own examination times are used, so it is much cheaper.

```{r grid}
fit <- pcdreg(pcd(id, tstart, tstop, count) ~ x1 + x2, data = d)
c(subjects = fit$n, examinations = fit$nexam,
  grid_times = fit$ngrid, events = fit$nevent)
```

Ties across subjects shrink the grid, and the grid drives the cost of the rate
model. Examination times recorded to the day or the week tie naturally; times
carried at full numerical precision never do, and the grid then grows with the
sample size. Rounding to the precision the visits were actually scheduled at is
worth doing.

# Checking before you fit

The quantities above are the quickest sanity check: `nexam` should match the
number of examinations you think you have, and `nevent` the total number of
events. If `nevent` is far larger than expected, the usual explanation is
cumulative counts.

```{r sanity}
c(rows_supplied = nrow(d),
  examination_rows = sum(!is.na(d$count)),
  events = sum(d$count, na.rm = TRUE))
```

# See also

`vignette("pcdreg")` for the models themselves, the choice of covariance
estimator, and the comparison between the rate and means models.
