---
title: "Regression analysis of panel count data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression analysis of panel count data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

```{r setup}
library(pcdreg)
```

## The data and the model

Panel count data arise when a recurrent event process is followed by periodic
examinations. At each visit you learn how many events have happened since the
last visit, but not when any of them happened. Hospitalisations counted at
quarterly clinic visits, tumours counted at scheduled sacrifices, and insurance
claims tallied at renewal all have this shape.

Write $N_i(t)$ for the number of events subject $i$ has experienced by time $t$,
observed at times $T_{i1} < \dots < T_{iJ_i}$, and $\Delta N_{ij}$ for the count
in $(T_{i,j-1}, T_{ij}]$. `panelrate()` fits the proportional rate model

$$E[dN(t) \mid X(t)] = \exp(\beta' X(t)) \, d\Lambda(t),$$

where $\Lambda$ is an unspecified non-decreasing baseline cumulative rate and
$X(t)$ may vary over time.

It is worth being clear about why the rate model rather than the proportional
*means* model $E[N(t) \mid X(t)] = \mu(t)\exp(\beta'X(t))$. The means model
requires $\mu$ to be non-decreasing, and when covariates fluctuate over time
that is hard to arrange and harder to interpret: nothing stops a fitted means
model from predicting a mean count that goes down. The rate model constrains
only $\exp(\beta'X(t))\,d\Lambda(t)$ to be positive, so predicted means increase
automatically, and $\exp(\beta'X(t))$ carries the same instantaneous
relative-risk reading that a hazard ratio has in survival analysis.

## Getting data into shape

The response is built by `PanelCount()`, which plays the role `Surv()` plays in
survival analysis and, like it, has two forms.

When covariates do not change over time, one row per examination is enough:

```{r three-arg}
visits <- data.frame(
  id    = c(1, 1, 1, 2, 2),
  time  = c(0.5, 1.1, 1.9, 0.7, 1.4),
  count = c(2, 0, 3, 1, 4),
  dose  = c(10, 10, 10, 25, 25)
)
PanelCount(visits$id, visits$time, visits$count)
```

Interval starts are filled in from the previous examination of the same subject,
with follow-up beginning at zero.

When covariates *do* change over time you need the counting process form, one
row per interval over which the covariates are constant. This is the same layout
`coxph()` uses for time-dependent covariates. A row whose `count` is a number
records an examination at `tstop`; a row whose `count` is `NA` records only that
a covariate changed there.

```{r four-arg}
cp <- data.frame(
  id     = c(1, 1, 1, 2),
  tstart = c(0.0, 0.5, 0.9, 0.0),
  tstop  = c(0.5, 0.9, 1.2, 0.8),
  count  = c(2, NA, 0, 3),
  dose   = c(10, 10, 25, 25)
)
PanelCount(cp$id, cp$tstart, cp$tstop, cp$count)
```

Subject 1 was examined at 0.5 and 1.2 with two events in the first interval and
none in the second; the dose changed at 0.9, part way through the second
interval, which is exactly the information the rate model can use and the means
model cannot.

Within a subject the intervals must be contiguous, start at zero, and end at an
examination. `panelrate()` checks all of this and says which subject is at fault
when it fails.

### Converting data held in the older layout

A common alternative layout repeats each panel count on every row of its
interval and marks the examinations with a separate indicator. Turning that into
the layout above takes one line, because a repeated count on a non-examination
row carries no information that the count on the examination row does not:

```{r convert, eval = FALSE}
new <- old
new$count[old$event == 0] <- NA
panelrate(PanelCount(id, tstart, tstop, count) ~ x1 + x2, data = new)
```

## Fitting

```{r fit}
set.seed(1)
d <- r_panel_count(200, beta = c(1, -1), lambda = function(t) 8 / (1 + t))
fit <- panelrate(PanelCount(id, tstart, tstop, count) ~ x1 + x2, data = d)
summary(fit)
```

There is no intercept: a constant in $X$ would be indistinguishable from a
rescaling of $\Lambda$, so `panelrate()` drops it and reports the baseline
separately.

Fitting maximises the likelihood implied by a nonhomogeneous Poisson process,
using an EM algorithm that treats the unobserved per-time counts as missing
data. That makes the awkward integral
$\int \exp(\beta'X(t)) d\Lambda(t)$ tractable and, usefully, means the
covariates are only ever needed at the observed examination times rather than as
complete trajectories.

The Poisson likelihood is a working device, not an assumption about the data.
The estimator stays consistent and asymptotically normal when the counts are not
Poisson at all.

## Which standard errors

That distinction matters most for the covariance. Three estimators are
available.

```{r variance}
se <- function(type) sqrt(diag(vcov(fit, type)))
rbind(robust = se("robust"), information = se("information"))
```

`"robust"`, the default, is the sandwich $\Omega^{-1} S \Omega^{-1}/n$ built from
the two matrices

```{r matrices}
fit$Omega
fit$S
```

both of which are by-products of the last EM iteration, so the robust standard
errors are essentially free. It is consistent whether or not the counts are
Poisson.

`"information"` is $S^{-1}/n$. Under the Poisson assumption $S$ is the efficient
information and this is the right answer; it is also a cheap stand-in for the
profile likelihood method, which it agrees with closely.

`"profile"` is the numerical profile likelihood estimator of Murphy and van der
Vaart. It is included for comparison with the literature, and because it is
expensive it is computed only on request:

```{r profile}
pfit <- panelrate(PanelCount(id, tstart, tstop, count) ~ x1 + x2, data = d,
                  profile = TRUE)
rbind(information = sqrt(diag(vcov(pfit, "information"))),
      profile     = sqrt(diag(vcov(pfit, "profile"))))
```

The two track each other closely, as the theory says they should, since both
estimate $S$. They are not identical: the profile version rests on a numerical
derivative with a step of order $n^{-1/2}$, so it carries a discretisation error
that shrinks only as the sample grows. That is the practical argument for
preferring `"information"` whenever the Poisson assumption is tenable, and it is
why `"profile"` is here mainly so that results can be compared with the
literature.

### What happens when the Poisson assumption fails

Real recurrent event data are usually overdispersed: some subjects are simply
more event-prone than their covariates suggest. Simulating that with a gamma
frailty leaves the rate model intact but destroys the Poisson structure.

```{r overdispersed}
set.seed(2)
od <- r_panel_count(200, beta = c(1, -1), frailty = 1)
odfit <- panelrate(PanelCount(id, tstart, tstop, count) ~ x1 + x2, data = od)
rbind(robust = sqrt(diag(vcov(odfit, "robust"))),
      information = sqrt(diag(vcov(odfit, "information"))))
```

The information-based standard errors are several times too small. The
simulations in the paper put the resulting coverage of nominal 95% intervals
around 15--22%. This is the reason `"robust"` is the default, and the reason to
think hard before reporting anything else.

## The baseline and predicted means

`baseline()` returns the estimated jumps and the cumulative baseline rate.

```{r baseline}
head(baseline(fit))
plot(fit)
lines(fit$baseline$time, 8 * log(1 + fit$baseline$time), col = 2, lty = 2)
```

The dashed line is the truth, $\Lambda(t) = 8\log(1+t)$.

Predicted mean counts follow each subject's covariate trajectory,
$E[N(t) \mid X] = \int_0^t \exp(\beta'X(s)) \, d\hat\Lambda(s)$:

```{r predict}
pred <- predict(fit, d, type = "mean")
head(pred)

one <- pred[pred$id == 1, ]
plot(one$time, one$mean, type = "s", xlab = "Time",
     ylab = "Predicted mean count", main = "Subject 1")
```

The curve is non-decreasing even though `x1` steps up or down part way through
follow-up, which is the structural advantage of the rate model.

## Computation

The EM algorithm converges linearly and, on this problem, slowly: plain EM can
still be moving in the fifth decimal place after tens of thousands of passes.
`panelrate()` therefore extrapolates the iterations using the SQUAREM scheme of
Varadhan and Roland, keeping an extrapolated point only when a stabilising EM
pass from it does at least as well on the observed data log likelihood. The
fixed point is unchanged; only the number of passes needed to reach it falls,
typically by an order of magnitude.

```{r passes}
c(iterations = fit$iterations, EM_passes = fit$passes,
  converged = fit$converged)
```

Set `accelerate = FALSE` in `panelrate_control()` to recover plain EM.

The dominant cost is the size of the pooled grid of distinct examination times,
because the covariate trajectory of every subject has to be evaluated at every
grid time within their follow-up. Examination times that tie across subjects,
which is what scheduled visits produce in practice, keep that grid small.

## Reference

Sun, D., Guo, Y., Li, Y., Tu, W. and Sun, J. (2024). A robust approach for
regression analysis of panel count data. *Bernoulli* **30**(4), 3251--3275.
[doi:10.3150/23-BEJ1713](https://doi.org/10.3150/23-BEJ1713)
