---
title: "Regression analysis of panel count data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Regression analysis of panel count data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 4)
```

```{r setup}
library(pcdreg)
```

## The data and the model

Panel count data arise when a recurrent event process is observed at periodic
examinations. Each visit gives the number of events since the previous visit,
but not their event times. Examples include hospitalisations counted at
quarterly clinic visits, tumours counted at scheduled sacrifices, and insurance
claims tallied at renewal.

Write $N_i(t)$ for the number of events subject $i$ has experienced by time $t$,
observed at times $T_{i1} < \dots < T_{iJ_i}$, and $\Delta N_{ij}$ for the count
in $(T_{i,j-1}, T_{ij}]$. `pcdreg()` fits the proportional rate model

$$E[dN(t) \mid X(t)] = \exp(\beta' X(t)) \, d\Lambda(t),$$

where $\Lambda$ is an unspecified non-decreasing baseline cumulative rate and
$X(t)$ may vary over time.

The proportional *means* model is $E[N(t) \mid X(t)] = \mu(t)\exp(\beta'X(t))$.
It requires $\mu$ to be non-decreasing. With covariates that fluctuate over
time, this condition is difficult to arrange and interpret: a fitted means
model can predict a declining mean count. The rate model requires only
$\exp(\beta'X(t))\,d\Lambda(t)$ to be positive. Predicted means therefore
increase automatically, and $\exp(\beta'X(t))$ has the instantaneous
relative-risk interpretation of a hazard ratio in survival analysis.

The means model is the established alternative. The package also fits it with
`pcdreg(model = "mean")`. The last section compares the models.

## Getting data into shape

`pcd()` constructs the response. It plays the same role as `Surv()` in survival
analysis and has two forms.

When covariates do not change over time, one row per examination is enough:

```{r three-arg}
visits <- data.frame(
  id    = c(1, 1, 1, 2, 2),
  time  = c(0.5, 1.1, 1.9, 0.7, 1.4),
  count = c(2, 0, 3, 1, 4),
  dose  = c(10, 10, 10, 25, 25)
)
pcd(visits$id, visits$time, visits$count)
```

Interval starts are filled in from the previous examination of the same subject,
with follow-up beginning at zero.

When covariates *do* change over time, use the counting process form: one row
per interval with constant covariates. This is the layout `coxph()` uses for
time-dependent covariates. A numeric `count` records an examination at `tstop`.
An `NA` `count` records a covariate change only.

```{r four-arg}
cp <- data.frame(
  id     = c(1, 1, 1, 2),
  tstart = c(0.0, 0.5, 0.9, 0.0),
  tstop  = c(0.5, 0.9, 1.2, 0.8),
  count  = c(2, NA, 0, 3),
  dose   = c(10, 10, 25, 25)
)
pcd(cp$id, cp$tstart, cp$tstop, cp$count)
```

Subject 1 was examined at 0.5 and 1.2, with two events in the first interval
and none in the second. The dose changed at 0.9, within the second interval.
The rate model uses this information; the means model cannot.

Within a subject the intervals must be contiguous, start at zero, and end at an
examination. `pcdreg()` checks all of this and says which subject is at fault
when it fails.

### Converting data held in the older layout

An older layout repeats each panel count on every row of its interval and marks
examinations with a separate indicator. Convert it in one line. A repeated count
on a non-examination row carries no information beyond the examination row:

```{r convert, eval = FALSE}
new <- old
new$count[old$event == 0] <- NA
pcdreg(pcd(id, tstart, tstop, count) ~ x1 + x2, data = new)
```

## Fitting

```{r fit}
set.seed(1)
d <- sim_pcd(200, beta = c(1, -1), lambda = function(t) 8 / (1 + t))
fit <- pcdreg(pcd(id, tstart, tstop, count) ~ x1 + x2, data = d)
summary(fit)
```

There is no intercept: a constant in $X$ would be indistinguishable from a
rescaling of $\Lambda$, so `pcdreg()` drops it and reports the baseline
separately.

Fitting maximises the likelihood implied by a nonhomogeneous Poisson process.
The EM algorithm treats the unobserved per-time counts as missing data. This
makes the integral $\int \exp(\beta'X(t)) d\Lambda(t)$ tractable and requires
covariates only at observed examination times, not as complete trajectories.

The Poisson likelihood is a working device, not an assumption about the data.
The estimator stays consistent and asymptotically normal when the counts are not
Poisson at all.

## Which standard errors

This distinction matters for the covariance. Three estimators are available.

```{r variance}
se <- function(type) sqrt(diag(vcov(fit, type)))
rbind(robust = se("robust"), information = se("information"))
```

`"robust"`, the default, is the sandwich $\Omega^{-1} S \Omega^{-1}/n$ built from
the two matrices

```{r matrices}
fit$Omega
fit$S
```

Both matrices are by-products of the last EM iteration, so the robust standard
errors are essentially free. The estimator is consistent whether or not the
counts are Poisson.

`"information"` is $S^{-1}/n$. Under the Poisson assumption, $S$ is the
efficient information and this is the right answer. It is also a cheap stand-in
for the profile likelihood method, with which it agrees closely.

`"profile"` is the numerical profile likelihood estimator of Murphy and van der
Vaart. It is included for comparison with the literature and is computed only
on request because it is expensive:

```{r profile}
pfit <- pcdreg(pcd(id, tstart, tstop, count) ~ x1 + x2, data = d,
                  profile = TRUE)
rbind(information = sqrt(diag(vcov(pfit, "information"))),
      profile     = sqrt(diag(vcov(pfit, "profile"))))
```

The two estimates track each other closely, as theory predicts, because both
estimate $S$. They differ because the profile version uses a numerical
derivative with a step of order $n^{-1/2}$ and therefore has a discretisation
error that shrinks only with sample size. Prefer `"information"` when the
Poisson assumption is tenable. `"profile"` is mainly provided for comparison
with the literature.

### What happens when the Poisson assumption fails

Real recurrent event data are usually overdispersed: some subjects are more
event-prone than their covariates suggest. A gamma frailty preserves the rate
model but destroys the Poisson structure.

```{r overdispersed}
set.seed(2)
od <- sim_pcd(200, beta = c(1, -1), frailty = 1)
odfit <- pcdreg(pcd(id, tstart, tstop, count) ~ x1 + x2, data = od)
rbind(robust = sqrt(diag(vcov(odfit, "robust"))),
      information = sqrt(diag(vcov(odfit, "information"))))
```

The information-based standard errors are several times too small. The paper's
simulations put coverage of the resulting nominal 95% intervals around 15--22%.
For this reason, `"robust"` is the default and other choices require care.

## The baseline and predicted means

`baseline()` returns the estimated jumps and the cumulative baseline rate.

```{r baseline, fig.alt = "Estimated cumulative baseline rate rising with time, with the true curve overlaid as a dashed red line and closely tracking it."}
head(baseline(fit))

truth <- data.frame(time = fit$baseline$time,
                    value = 8 * log(1 + fit$baseline$time))
autoplot(fit) +
  ggplot2::geom_line(data = truth, colour = "#c0392b", linetype = 2)
```

The dashed line is the truth, $\Lambda(t) = 8\log(1+t)$.

Predicted mean counts follow each subject's covariate trajectory,
$E[N(t) \mid X] = \int_0^t \exp(\beta'X(s)) \, d\hat\Lambda(s)$:

```{r predict, fig.alt = "Predicted mean count for one subject as a step function, rising monotonically over follow-up."}
pred <- predict(fit, d, type = "mean")
head(pred)

one <- pred[pred$id == 1, ]
ggplot2::ggplot(one, ggplot2::aes(x = time, y = mean)) +
  ggplot2::geom_step(colour = "#2a78d6", linewidth = 0.6) +
  ggplot2::labs(x = "Time", y = "Predicted mean count", title = "Subject 1") +
  ggplot2::theme_minimal(base_size = 11)
```

The curve is non-decreasing even though `x1` changes part way through follow-up.
This is a structural advantage of the rate model.

## Tidy output

`tidy()`, `glance()` and `augment()` methods are provided for the broom
generics.

```{r tidiers}
tidy(fit, conf.int = TRUE)
glance(fit)
```

`exponentiate = TRUE` reports $e^{\beta}$, the multiplicative effect on the
rate, read as a hazard ratio is read.

```{r exponentiate}
tidy(fit, conf.int = TRUE, exponentiate = TRUE)
```

`augment()` attaches the fitted mean to each row, together with the observed
cumulative count and their difference. `.resid` is the excess of events seen so
far over the number the model implies, so it is the quantity to plot against
time or against a covariate when checking fit.

```{r augment}
aug <- augment(fit)
head(aug[!is.na(aug$.observed),
         c("id", "tstop", ".fitted", ".observed", ".resid")])
```

## The means model

`pcdreg(model = "mean")` fits the other classical model for these data,

$$E[N(t) \mid X(t)] = \mu(t)\exp(\beta'X(t)),$$

by the estimating equation of Hu, Sun and Wei (2003). It is the comparator in
the paper's application. It is inexpensive because only examination times enter;
the grid expansion required by the rate model is unnecessary.

```{r meanmodel}
mfit <- pcdreg(pcd(id, tstart, tstop, count) ~ x1 + x2, data = d, model = "mean")
summary(mfit)
```

There is no likelihood behind this estimator, so there is no log likelihood to
report and `vcov()` offers only the sandwich.

The two fits require care in interpretation.

```{r compare}
cbind(rate = coef(fit), mean = coef(mfit))
```

These numbers do not estimate one quantity. With time-varying covariates, the
models are not reparametrisations of each other: $\beta$ acts on the
instantaneous rate in one and the cumulative mean in the other. They answer
different questions and need not agree. These data were generated from the rate
model, so `pcdreg()` recovers the simulation values and `pcdreg(model = "mean")`
does not.

Second, nothing constrains the fitted $\hat\mu$ to increase:

```{r monotone}
mu <- baseline(mfit)$mean
c(increasing = !is.unsorted(mu), decreases_at = sum(diff(mu) < 0))
```

This is the difficulty that the package's rate model addresses. It appears in
ordinary simulated data. A declining mean function is not a numerical failure.
It is the estimator's result when covariate values move up and down. Predicted
means from `pcdreg(model = "mean")` can therefore decrease, while those from `pcdreg()`
cannot.

## Computation

The EM algorithm converges linearly and slowly on this problem: plain EM can
still be changing in the fifth decimal place after tens of thousands of passes.
`pcdreg()` extrapolates the iterations with the SQUAREM scheme of Varadhan and
Roland. An extrapolated point is kept only if a stabilising EM pass from it does
at least as well on the observed-data log likelihood. The fixed point is
unchanged. The required number of passes typically falls by an order of
magnitude.

```{r passes}
c(iterations = fit$iterations, EM_passes = fit$passes,
  converged = fit$converged)
```

Set `accelerate = FALSE` in `pcdreg_control()` to recover plain EM.

The dominant cost is the pooled grid of distinct examination times. The
covariate trajectory for every subject is evaluated at every grid time within
follow-up. Tied examination times across subjects, as with scheduled visits,
keep the grid small.

## References

Sun, D., Guo, Y., Li, Y., Tu, W. and Sun, J. (2024). A robust approach for
regression analysis of panel count data. *Bernoulli* **30**(4), 3251--3275.
[doi:10.3150/23-BEJ1713](https://doi.org/10.3150/23-BEJ1713)

Hu, X. J., Sun, J. and Wei, L. J. (2003). Regression parameter estimation from
panel counts. *Scandinavian Journal of Statistics* **30**(1), 25--43.
[doi:10.1111/1467-9469.00316](https://doi.org/10.1111/1467-9469.00316)

Varadhan, R. and Roland, C. (2008). Simple and globally convergent methods for
accelerating the convergence of any EM algorithm. *Scandinavian Journal of
Statistics* **35**, 335--353.
